\documentclass[a4paper]{ltxdoc}
\usepackage[a4paper,left=2.25cm,right=2.25cm,top=2.5cm,bottom=2.5cm,nohead]{geometry}
\usepackage{amsmath}
\usepackage{makeidx}
\usepackage{tikz}
\usepackage[T1]{fontenc}
\usepackage{trace}

% These are required to typeset the manual.
\usepackage{calc}
\usepackage{hyperref}

% We can't use \usetikzlibrary{spline} because TeX won't find the correct
% file one directory up. So fake it.
%\usetikzlibrary{spline}
\makeatletter
\input{../tikzlibraryspline.code}
\let\tikz@library@spline@loaded\empty
\makeatother

\input{pgfmanual-en-macros}
\makeindex

\title{\tikzname\ spline library}
\author{Stephen Checkoway}
\date{2018-03-13}

\begin{document}
\maketitle

\section{Constructing splines through points}

\begin{tikzlibrary}{spline}
This library enables drawing cubic B\'ezier splines through specified points.
\end{tikzlibrary}

When you load this library, the following options become available:

\begin{keylist}{%
    /tikz/spline through=\meta{coordinates or nodes},
    /tikz/closed spline through=\meta{coordinates or nodes}}
These options must be given to a |to| or |edge| path command. The
\meta{coordinates or nodes} should be a sequence of \tikzname\ coordinates or
node names, one directly after the other without commas (like with the
|plot coordinates| path operation). Examples are |(1,0) (2,2)| or
|(a) (1,0) (b)|, where |a| and |b| are nodes. For |closed spline through|, the
\meta{coordinates or nodes} must contain at least two coordinates or node
names.

A spline is constructed from the current position through each of the
\meta{coordinates or nodes} and ending either at the target of the |to| or |edge|
path command for |spline through| or at the current position for 
|closed spline through|.
\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw=red, circle] (A) at (1,0) {A};
  \node[draw=blue, rectangle] (B) at (2,2) {B};
  \node[draw=green!60!black, circle] (C) at (3,1) {C};
  
  \draw (A) to[spline through={(0,1)(B)(2,1)}] (C);
\end{tikzpicture}
\end{codeexample}

\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw=red, circle] (A) at (1,0) {A};
  \node[draw=blue, rectangle] (B) at (2,2) {B};
  \node[draw=green!60!black, circle] (C) at (3,1) {C};
  
  \draw (A) to[closed spline through={(0,1)(B)(2,1)(C)}] (A);
\end{tikzpicture}
\end{codeexample}

If you want the curve to not stop at the border of a node, the |center| anchor
(or any other node anchor) can be given.
\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw=red, circle] (A) at (1,0) {A};
  \node[draw=blue, rectangle] (B) at (2,2) {B};
  \node[draw=green!60!black, circle] (C) at (3,1) {C};
  
  \draw (A.135) to[spline through={(0,1)(B.south)(2,1)}] (C.center);
\end{tikzpicture}
\end{codeexample}

\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw=red, circle] (A) at (1,0) {A};
  \node[draw=blue, rectangle] (B) at (2,2) {B};
  \node[draw=green!60!black, circle] (C) at (3,1) {C};
  
  \draw (A.135) to[closed spline through={(0,1)(B.south)(2,1)(C.center)}] (A);
\end{tikzpicture}
\end{codeexample}
\end{keylist}

\begin{key}{/tikz/spline coordinates=\meta{name prefix} (default spline)}
This option must be given to a |to| or |edge| path command that is also given
one of |spline through| or |closed spline through|.

The spline's ``knot'' points---the specified points of the spline---and
control points are made available as coordinates. If the spline has $n$
segments, then the knot points $K_1,K_2,\dotsc,K_{n+1}$ have names \meta{name prefix}
|K-1|, \meta{name prefix} |K-2|, and so on.
(For a closed spline, the $(n+1)$th knot point is the starting
point of the spline.) The $2n$ control points $P_1,P_2,\dotsc,P_n$ and
$Q_1,Q_2,\dotsc,Q_n$ 
are available with names \meta{name prefix} |P-1|, \meta{name prefix} |Q-1|,
\meta{name prefix} |P-2|, \meta{name prefix} |Q-2|, and so on.

The $i$th segment of the spline is a cubic B\`ezier curve from $K_i$ to
$K_{i+1}$ with control points $P_i$ and $Q_i$. In \tikzname, the third segment
can be expressed as\\[.5\baselineskip]
\begingroup
\frenchspacing
\null\quad |(spline K-3) .. controls (spline P-3) and (spline Q-3) .. (spline K-4)|
\endgroup

The number of segments is available as the \TeX\ macro
\cs{tikzsplinesegments}.

\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \coordinate (A) at (1,0);
  \coordinate (B) at (2,2);
  \coordinate (C) at (3,1);
  
  \draw (A) to[spline coordinates=S,
               closed spline through={(0,1)(B)(2,1)(C)}] (A);
  \foreach \i [evaluate=\i as \j using \i+1] in {1,2,...,\tikzsplinesegments} {
    \draw[semithick]  (S K-\i) -- (S P-\i)
                      (S K-\j) -- (S Q-\i);
    \path[fill=green!60!black] (S P-\i) circle (2pt);
    \path[fill=red] (S Q-\i) circle (2pt);
  }
\end{tikzpicture}
\end{codeexample}
\end{key}

\section{Intersection with shapes}
When \tikzname\ creates a B\'ezier curve where one or both of the end points
is a node, it constructs the line through the center of the node tangent to
the curve at that point. \tikzname\ then computes the intersection of the line
with the node's background path. Finally, it adjusts the end point of the
curve to be the intersection point. This happens without changing the curve's
control points. The result is a completely different curve. Compare the red
and black curves.
\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw, circle] (A) at (0,1) {A};
  \node[draw] (B) at (3,1) {B};
  \draw[red] (A.center) .. controls +(0,1) and +(0,1) .. (B.center);
  \draw (A) .. controls +(0,1) and +(0,1) .. (B);
\end{tikzpicture}
\end{codeexample}

Normally, this difference is not a big deal and it's fast for \pgfname\ to
compute this border point. Unfortunately, when the same procedure is applied
to a B\`ezier spline, the result is slightly unsettling.

In the example below, the red line has been drawn explicitly from the
|closed spline through|'s knot and control points whereas the black line uses
the same control points but the start and end points have been modified
according to \tikzname's normal procedure.
\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,3);
  \node[draw,circle]    (A) at (0,1)   {A};
  \node[draw]           (B) at (3,1)   {B};
  \node[draw,circle]    (C) at (2,3)   {C};
  \node[draw,circle]    (D) at (1.5,0) {D};
  
  \path
    (A.center) to[spline coordinates=S,
                  closed spline through=(B.center)(C.center)(D.center)] (A);
  \draw[red]
    (A.center) .. controls (S P-1) and (S Q-1) .. (B.center)
               .. controls (S P-2) and (S Q-2) .. (C.center)
               .. controls (S P-3) and (S Q-3) .. (D.center)
               .. controls (S P-4) and (S Q-4) .. (A.center);
  \draw (A) .. controls (S P-1) and (S Q-1) .. (B)
            .. controls (S P-2) and (S Q-2) .. (C)
            .. controls (S P-3) and (S Q-3) .. (D)
            .. controls (S P-4) and (S Q-4) .. (A);
\end{tikzpicture}
\end{codeexample}
None of the black line's intersections with the node borders look quite right.

The |spline| library uses the \pgfname\ |intersections| library to compute the
intersection of the spline with node background shapes. This is slower, but
the result is significantly better.

\begin{codeexample}[]
\begin{tikzpicture}
  \draw[help lines] (0,0) grid (3,3);
  \node[draw,circle]    (A) at (0,1)   {A};
  \node[draw]           (B) at (3,1)   {B};
  \node[draw,circle]    (C) at (2,3)   {C};
  \node[draw,circle]    (D) at (1.5,0) {D};
  
  \path
    (A.center) to[spline coordinates=S,
                  closed spline through=(B.center)(C.center)(D.center)] (A);
  \draw[red]
    (A.center) .. controls (S P-1) and (S Q-1) .. (B.center)
               .. controls (S P-2) and (S Q-2) .. (C.center)
               .. controls (S P-3) and (S Q-3) .. (D.center)
               .. controls (S P-4) and (S Q-4) .. (A.center);
  \draw[thick] (A) to[closed spline through=(B)(C)(D)] (A);
\end{tikzpicture}
\end{codeexample}

\section{Placing nodes along a spline}
Nodes can be placed along a spline as with any other standard path by giving a
|pos=|\meta{time} option to the node.
\begin{codeexample}[]
\begin{tikzpicture}[thick]
  \draw[help lines] (0,0) grid (3,2);
  \node[draw=red, circle] (A) at (1,0) {A};
  \node[draw=blue, rectangle] (B) at (2,2) {B};
  \node[draw=green!60!black, circle] (C) at (3,1) {C};
  
  \draw (A) to[spline through={(0,1)(B)(2,1)}, node font=\footnotesize]
    node foreach \pos in {0, .25, .5, .75, 1}
      [pos=\pos, fill=blue, circle, inner sep=1pt] {}
    node[pos=0, auto] {$0$}
    node[pos=.25, auto] {$.25$}
    node[pos=.5, auto] {$.5$}
    node[pos=.75, auto, swap] {$.75$}
    node[pos=1, auto] {$1$}
    (C);
\end{tikzpicture}
\end{codeexample}
Note that the distance between |pos=.25| and |pos=.5| is much greater than the
distance between |pos=.75| and |pos=1|.

The position along the curve is computed by first dividing $[0,1]$ into $n$
equal-size intervals for a spline with $n$ segments. After selecting the
segment, the position along the segment is computed in the normal fashion for
curves with one exception. If the start of the spline is a node with a
background shape, |pos=0| is set to the intersection of the curve and the
shape. Similarly, if the end of the spline is a node with a background shape,
|pos=1| is set to the intersection of the curve with the shape.

Just as with normal \tikzname\ curves, positioning nodes requires a bit of a
trial-and-error.

\section{Debugging constructed splines}
\begin{key}{/tikz/spline debug=\meta{boolean} (default true, initially false)}
When given as an option to a |to| or |edge| path command, this causes
debugging output to be written to the log and the terminal. It is not intended
to be useful to end users but the author found it invaluable for debugging the
code.
\end{key}

\printindex
\end{document}
% vim: set sw=2 sts=2 ts=8 et:
